Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-03-10 ThesisJP/Background/background.tex
Medical ultrasound imaging.

\section {Medical ultrasound imaging}\label{sec:ultrasound}
Ultrasound imaging encompasses technology which generates images based on sound whose frequencies we can not hear. For medical ultrasound imaging frequencies between 2 MHz and 10 MHz are typically used. Which frequency to select depends on how deep into the body we want to image, this because human tissue exhibits frequency dependent attenuation. Frequency dependent attenuation is the reason why a frequency around 3 MHz is used for cardiac imaging and a frequency around 7 MHz is used for more shallow imaging (e.g. vascular imaging). To generate the ultrasonic frequencies a piezoelectric ceramic is used. A piezoelectric material is special in that way that ultrasound vibrations are generated when voltage is applied and \textit{vice versa}. It can therefore be used to both send and receive ultrasound signals, like a combined microphone and speaker. Piezoelectric elements organized in an array, is referred to as an ultrasound \textit{probe}, and a special probe is typically designed for each medical imaging modality. For cardiac ultrasound imaging the probe must not only be designed for a certain frequency range, its footprint also needs to be smaller than the intercostal window in order to avoid shadowing by ribs. A line in an ultrasound image is generated by firing a ultrasound pulse into the body and sample the received echo. This amplitude trace is known as an A-mode image.

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed or sensed equally in all directions. When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse in space integrated across time.} of ultrasound by diffraction as the size of the array  (the \textit{aperture}) grows. The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:
\begin{align}\label{eq:res}
\delta\theta_a = \frac{\lambda}{D} = \frac{c}{fD},
\end{align}
\nomi{$\delta\theta_a$}{Angular resolution of an aperture}where \nom{$\lambda$}{Wavelength} and \nom{$f$}{Frequency} is the ultrasound wavelength and frequency respectively, $c$ is the speed of sound, and \nom{$D$}{Aperture size} is the aperture width. With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$, the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree, which at 8 cm gives a lateral resolution of around 2 mm. The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution.

@@

If time delays are applied on the signal sent to the different elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das}). This technique is known as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses. For medical ultrasound imaging there are two main types of ultrasound probes, linear and phased probes \cite{AngelUltrasound}. A cardiac probe is a phased array, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$ radians, resulting in an imaging sector which are $\theta$ radians wide. This kind of image, where A-mode images are acquired from multiple spatial locations in the $xz$-plane, is known as a B-mode image (B for brightness). If A-mode images are steered in the $yz$-plane as well we get what is known as a volumetric B-mode or 3D ultrasound image.

If \nom{$M$}{Number of array elements} is the number of elements in the array, and \nom{$\Delta_m$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, the standard form of beamforming used on receive, known as \textit{delay-and-sum}, is given as:
\begin{align}\label{eq:das}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
\end{align}
where \nom{$\vec{w}$}{Array weight vector (apodization)} is know as the array \textit{weight vector} or \textit{apodization}, and $z[n]$ is the output A-mode image in the direction given by $\Delta_m[n]$.  Delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das}. If $p_o$ is the phase center of our array and $p$ is the point where we want to steer and focus the ultrasound beam, we have to use the following delays:
\begin{align}
\Delta_m[n] =\frac{|p_o+p| - |p_m+p|}{c},
\end{align}
where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element. On transmit the inverse set of delays has to be used. Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse, transmit delays therefore do not vary within an A-mode image.

When generating a B-mode image we have to insure that the spatial and temporal sampling is sufficient. We have already seen that the lateral resolution of an ultrasound imaging probe is given by (\ref{eq:res}).  Hergum \textit{et al.} \cite{Hergum2009} have further shown that the lateral frequency support is governed by the sum of the transmit (\nom{tx}{Transmit}) and receive (\nom{rx}{Receive}) aperture. The lateral resolution of an active imaging system is then found to be
\begin{align}
\delta\theta = \frac{\lambda}{D_{tx} + D_{rx}}.
\end{align}
\nomi{$\delta\theta$}{Angular resolution of a two-way imaging system}\nomi{$D_{tx}$}{Aperture size on transmit}\nomi{$D_{rx}$}{Aperture size on receive}Then, in order to properly sample an imaging sector of $\theta$ radians we have to acquire 
\begin{align}
N_\theta = \frac{\theta}{\delta\theta}
\end{align}
\nomi{$N_\theta$}{Number of lines required for proper sampling of an imaging sector}A-mode images uniformly distributed across the sector. If \nom{$r$}{Imaging range.} is the maximum range we want to image we get an imaging frame rate of
\begin{align}
FR = \frac{c}{2rN_\theta}
\end{align} 
frames per second (\nom{FPS}{Frames per second}). With our 2 cm cardiac probe from the earlier example we get a frame rate of around $55$ FPS when imaging a 70\degree{} sector with $r=15$ cm.

The required axial sampling is dictated by the transmit frequency and the frequency bandwidth of the probe. From Fourier theory we know that a short impulse consist of many frequency components. A short ultrasound pulse, which would yield high axial resolution, therefore requires a large frequency bandwidth. A ultrasound probe has normally a relative bandwidth between 50 and 100 \%. This corresponds to transmitting a pulse which are between 2 and 1 oscillation long. In the front-end of an ultrasound imaging system a high sampling frequency is typically used (e.g. 40 MHz). However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \textit{\nom{IQ}{In-phase quadrature (ultrasound data format).} demodulation}. In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation. Number of samples in one A-mode image is then found to be:
\begin{align}
N = \frac{2r}{c}B.
\end{align}
\nomi{$N$}{Number of samples in range}where \nom{$B$}{Frequency bandwidth} is the bandwidth, $B = B_wf_c$ (where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth and \nom{$f_c$}{Center frequency of ultrasound pulse} is the center frequency of the transmitted pulse). The total number of samples in an B-mode image is then 
\begin{align}
N_{\text{B-mode}} = N_\theta N.
\end{align}
For our cardiac probe has a relative bandwidth of $80$ \% we have to process almost $50000$ $M$-long vectors of element data every $1/55$ second.

@@

Medical ultrasound imaging is considered to be a low-risk tool for non-invasive investigation. There are no gamma rays or high intensity magnetic fields involved, just sound pressure below well controlled limits. It is also far less costly than other imaging modalities and offers real-time inspection of the human body.  Ultrasound imaging is well know for its use in fetal imaging during pregnancy, however, in this thesis we will focus on different modality, namely cardiac ultrasound imaging. An illustration of the human heart is therefore included in Fig.\,\ref{fig:human_heart}. Throughout the thesis we will refer to some of the parts shown in Fig.\,\ref{fig:human_heart}. Especially note the four chambers of the heart, the left and right atrium, and the left and right ventricle. The tissue separating the left and right ventricle is know as the inter-ventricular septum. The left side is the strongest and hardest working part of the heart, and it is responsible of pumping blood out to the whole body through the aortic valve. When blood flows out the aorta it is important that the mitral valve, that separates the left atrium and ventricle, is closed and does not leak. The left ventricle and the mitral valve therefore receives special attention during cardiac ultrasound examinations. When visualizing 3D ultrasound images we will have a strong focus on visibility of cardiac tissue, and note for later that the endocardium refers to the innermost layer of the cardiac muscle.

>>>>

\section {Medical ultrasound imaging}\label{sec:ultrasound}
Ultrasound imaging encompasses technology which generates images based on sound whose frequencies we can not hear. For medical ultrasound imaging frequencies between 2 MHz and 10 MHz are typically used. Which frequency to select depends on how deep into the body we want to image, this because human tissue exhibits frequency dependent attenuation. Frequency dependent attenuation is the reason why a frequency around 3 MHz is used for fetal and cardiac imaging and a frequency around 7 MHz is used for more shallow imaging (e.g. vascular imaging).

Medical ultrasound imaging is considered to be a low-risk tool for non-invasive investigation. There are no gamma rays or high intensity magnetic fields involved, just sound pressure below well controlled limits. It is also far less costly than other imaging modalities and offers real-time inspection of the human body.  Ultrasound imaging is well know for its use in fetal imaging during pregnancy, however, in this thesis we will focus on a different modality, namely cardiac ultrasound imaging. An illustration of the human heart is therefore included in Fig.\,\ref{fig:human_heart}. Throughout the thesis we will refer to some of the parts shown in Fig.\,\ref{fig:human_heart}. Especially note the four chambers of the heart, the left and right atrium, and the left and right ventricle. The tissue separating the left and right ventricle is know as the inter-ventricular septum. The left side is the strongest and hardest working part of the heart, and it is responsible of pumping blood out to the whole body through the aortic valve. When blood flows out the aorta it is important that the mitral valve, that separates the left atrium and ventricle, is closed and does not leak. The left ventricle and the mitral valve therefore receives special attention during cardiac ultrasound examinations. When visualizing 3D ultrasound images in \textbf{Paper IV} we will have a strong focus on visibility of cardiac tissue, thus note for later that the endocardium refers to the innermost layer of the cardiac muscle.

To generate the ultrasonic frequencies used for imaging a piezoelectric ceramic is used. A piezoelectric material is special in that way that ultrasound vibrations are generated when voltage is applied and \textit{vice versa}. It can therefore be used to both send and receive ultrasound signals, like a combined microphone and speaker. Piezoelectric elements organized in an array, is referred to as an ultrasound \textit{probe}, and a special probe is typically designed for each medical imaging modality. For cardiac ultrasound imaging the probe must not only be designed for a certain frequency range, its footprint also needs to be smaller than the intercostal window in order to avoid shadowing by ribs. A line in an ultrasound image is generated by firing a ultrasound pulse into the body and sample the received echo. This amplitude trace is known as an A-mode image.

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed or sensed equally in all directions. When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse in space integrated across time.} of ultrasound by diffraction as the size of the array  (the \textit{aperture}) grows. The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:
\begin{align}\label{eq:res}
\delta\theta_a = \frac{\lambda}{D} = \frac{c}{fD},
\end{align}
\nomi{$\delta\theta_a$}{Angular resolution of an aperture}where \nom{$\lambda$}{Wavelength} and \nom{$f$}{Frequency} is the ultrasound wavelength and frequency respectively, $c$ is the speed of sound, and \nom{$D$}{Aperture size} is the aperture width. With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$, the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree, which at 8 cm range gives a lateral resolution of around 2 mm. The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution.

@@

If time delays are applied on the signal sent to the different elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das}). This technique is known as \textit{beamforming}, and for ultrasound imaging it is traditionally applied both when transmitting and receiving ultrasound pulses. For medical ultrasound imaging there are two main types of ultrasound probes, linear and phased probes \cite{AngelUltrasound}. A cardiac probe is a phased probe, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$ radians, resulting in an imaging sector which are $\theta$ radians wide. This kind of image, where A-mode images are acquired from multiple spatial locations in the $xz$-plane, is known as a B-mode image (B for brightness). If A-mode images are steered in the $yz$-plane as well we get what is known as a volumetric B-mode or 3D ultrasound image.

If \nom{$M$}{Number of array elements} is the number of elements in the array, and \nom{$\Delta_m$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, the standard form of beamforming used on receive, known as \textit{delay-and-sum}, is given as:
\begin{align}\label{eq:das}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
\end{align}
where \nom{$\vec{w}$}{Array weight vector (apodization)} is know as the array \textit{weight vector} or \textit{apodization}, and $z[n]$ is the output A-mode image in the direction given by $\Delta_m[n]$.  A schematic overview of delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das}. If $p_o$ is the phase center of our array and $p$ is the point where we want to steer and focus the ultrasound beam, we have to use the following delays:
\begin{align}
\Delta_m[n] =\frac{|p_o+p| - |p_m+p|}{c},
\end{align}
where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element. On transmit the inverse set of delays has to be used. Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse, transmit delays therefore do not vary within an A-mode image.

When generating a B-mode image we have to insure that the spatial and temporal sampling are sufficient. We have already seen that the lateral resolution of an ultrasound probe is given by (\ref{eq:res}).  Hergum \textit{et al.} \cite{Hergum2009} have further shown that the lateral frequency support for a system which both emit and receive energy is governed by the sum of the transmit (\nom{tx}{Transmit}) and receive (\nom{rx}{Receive}) aperture. The lateral resolution of what is known as an active imaging system is found to be
\begin{align}
\delta\theta = \frac{\lambda}{D_{tx} + D_{rx}}.
\end{align}
\nomi{$\delta\theta$}{Angular resolution of a two-way imaging system}\nomi{$D_{tx}$}{Aperture size on transmit}\nomi{$D_{rx}$}{Aperture size on receive}Then, in order to properly sample an imaging sector of $\theta$ radians we have to acquire 
\begin{align}
N_\theta = \frac{\theta}{\delta\theta}
\end{align}
\nomi{$N_\theta$}{Number of lines required for proper sampling of an imaging sector}A-mode images uniformly distributed across the sector. If \nom{$r$}{Imaging range.} is the maximum range we want to image we get an imaging frame rate of
\begin{align}
f_{\text{FPS}} = \frac{c}{2rN_\theta}
\end{align} 
frames per second (\nom{FPS}{Frames per second}). With our 2 cm cardiac probe from the earlier example we get a frame rate of around $55$ FPS when imaging a 70\degree{} sector with $r=15$ cm.

The required axial sampling is dictated by the transmit frequency and the frequency bandwidth of the probe. From Fourier theory we know that a short impulse consist of many frequency components. A short ultrasound pulse, which would yield high axial resolution, therefore requires a large frequency bandwidth. A ultrasound probe has normally a relative bandwidth between 50 and 100 \%. This corresponds to transmitting a pulse which are between 2 and 1 oscillation long. In the front-end of an ultrasound imaging system a high sampling frequency is typically used (e.g. 40 MHz). However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \textit{\nom{IQ}{In-phase quadrature (ultrasound data format).} demodulation}. In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation. The number of samples in one A-mode image is after this process given by:
\begin{align}
N = \frac{2r}{c}B,
\end{align}
\nomi{$N$}{Number of samples in range}where \nom{$B$}{Frequency bandwidth} is the bandwidth, $B = B_wf_c$ (where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth and \nom{$f_c$}{Center frequency of ultrasound pulse} is the center frequency of the transmitted pulse). The total number of samples in an B-mode image is then 
\begin{align}
N_{\text{B-mode}} = N_\theta N.
\end{align}
If our cardiac probe has a relative bandwidth of $80$ \% we have to process almost $50000$ $M$-long vectors of element data every $1/55$ second. The formula for number of FLOPS needed for DAS beamforming is then given as:
\begin{align}
\text{FLOPS}_{\text{DAS}} = MN_{\text{B-mode}}f_{\text{FPS}}
\end{align}
If our cardiac probe has $M=128$ elements, DAS beamforming requires a processing throughput of around $350$ MFLOPS. Hence, far below the peak processing throughput of modern GPUs and CPUs.

@@
